Causal Bandits: Learning Good Interventions via Causal Inference

### 2016-10-28 causal-bandit/nips_2016/experiments.tex
mark's changes

In figure \ref{fig:simple_vs_m} we fix the number of variables $N$ and the horizon $T$ and compare the performance of the algorithms as $m$ increases. The regret for the Successive Reject algorithm is constant as it depends only on the reward distribution and has no knowledge of the causal structure. For the causal algorithms it increases approximately with $\sqrt{m}$. As $m$ approaches $N$, the gain the causal algorithms obtain from knowledge of the structure is outweighed by fact they do not leverage the observed rewards to focus sampling effort on actions with high pay-offs.

@@

Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\smash{\epsilon = \sqrt{N/(8T)}}$ , is just too small to be learned. This gap is learn-able by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$. In figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially. The decay constant is larger for the causal algorithms as they have observed a greater effective number of samples for a given $T$.

>>>>

In Figure \ref{fig:simple_vs_m} we fix the number of variables $N$ and the horizon $T$ and compare the performance of the algorithms as $m$ increases. The regret for the Successive Reject algorithm is constant as it depends only on the reward distribution and has no knowledge of the causal structure. For the causal algorithms it increases approximately with $\sqrt{m}$. As $m$ approaches $N$, the gain the causal algorithms obtain from knowledge of the structure is outweighed by fact they do not leverage the observed rewards to focus sampling effort on actions with high pay-offs.

@@

Figure \ref{fig:simple_vs_T_vary_epsilon} demonstrates the performance of the algorithms in the worst case environment for standard bandits, where the gap between the optimal and sub-optimal arms, $\smash{\epsilon = \sqrt{N/(8T)}}$ , is just too small to be learned. This gap is learnable by the causal algorithms, for which the worst case $\epsilon$ depends on $m \ll N$. In Figure \ref{fig:simple_vs_T} we fix $N$ and $\epsilon$ and observe that, for sufficiently large $T$, the regret decays exponentially. The decay constant is larger for the causal algorithms as they have observed a greater effective number of samples for a given $T$.

### 2016-10-28 causal-bandit/nips_2016/general.tex
updates additional experiments and proofs
mark's changes

We now consider the more general problem where the graph structure is known, but arbitrary. For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation). However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \cite{Pearl2000}. 
\eq{
\P{X_1...X_N|do(X_i=j)} = 
\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)\,, 
} 
where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function.

@@

To solve the general problem we need an estimator for each action that incorporates information obtained from every other action and a way to optimally allocate samples to actions. To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) are known. These could be estimated from experimental data on the same covariates but where the outcome of interest differed, such that $Y$ was not included, or similarly from observational data subject to identifiability constraints. Let $\eta$ be a distribution on available interventions $a \in \calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$. Define $Q = \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}$ to be the mixture distribution over the interventions with respect to $\eta$.

%we firstly note that the causal effect of an action $a$ on $Y$ can be factored into the causal effect of $a$ on the parents of $Y$ and the effect of the parents of $Y$ on $Y$. $\P{Y|a} = \P{Y|\parents{Y}}\P{\parents{Y}|a}$. We then

\begin{wrapfigure}[12]{r}{0.5\textwidth}
\vspace{-25pt}
\begin{minipage}{.5\textwidth}
\begin{algorithm}[H]
\caption{General Algorithm}\label{alg:general}
\begin{algorithmic}
\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
\FOR{$t \in \set{1,\ldots,T}$}
\STATE Sample action $a_t$ from $\eta$
\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
\ENDFOR
\FOR{$a \in \calA$}
\STATE
\eq {
\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}
}
\ENDFOR
\STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
\end{algorithmic}
\end{algorithm}
\end{minipage}
\end{wrapfigure}

@@

The second observation is that, in the parallel bandit setting, $m(\eta^*) \leq 2m(\boldsymbol{q})$. This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q})} / 2m(\boldsymbol{q})$ for the actions corresponding to $do(X_i=j)$, and applying an argument like that for Proposition~\ref{pro:m-bound}.

>>>>

We now consider the more general problem where the graph structure is known, but arbitrary. For general graphs, $\P{Y|X_i=j} \neq \P{Y|do(X_i=j)}$ (correlation is not causation). However, if all the variables are observable, any causal distribution $\P{X_1...X_N|do(X_i=j)}$ can be expressed in terms of observational distributions via the truncated factorization formula \citep{Pearl2000}. 
\eq{
\P{X_1...X_N|do(X_i=j)} = 
\prod_{k \neq i}\P{X_k|\parents{X_k}}\delta(X_i - j)\,, 
} 
where $\parents{X_k}$ denotes the parents of $X_k$ and $\delta$ is the dirac delta function.

@@

To solve the general problem we need an estimator for each action that incorporates information obtained from every other action and a way to optimally 
allocate samples to actions. To address this difficult problem, we assume the conditional interventional distributions $\P{\parents{Y}|a}$ (but not $\P{Y|a}$) 
are known. These could be estimated from experimental data on the same covariates but where the outcome of interest differed, such that $Y$ was not included, 
or similarly from observational data subject to identifiability constraints. Of course this is a somewhat limiting assumption, but seems like a natural place to
start. The challenge of estimating the conditional distributions for all variables in an optimal way is left as an interesting future direction.
Let $\eta$ be a distribution on available interventions $a \in \calA$ so $\eta_a \geq 0$ and $\sum_{a \in \calA} \eta_a = 1$. 
Define $Q = \sum_{a \in \calA} \eta_a \P{\parents{Y}|a}$ to be the mixture distribution over the interventions with respect to $\eta$.

\begin{wrapfigure}[11]{r}{0.5\textwidth}
\vspace{-25pt}
\begin{minipage}{.5\textwidth}
\begin{algorithm}[H]
\caption{General Algorithm}\label{alg:general}
\begin{algorithmic}
\STATE {\bf Input:} $T$, $\eta \in [0,1]^{\calA}$, $B \in [0,\infty)^{\calA}$
\FOR{$t \in \set{1,\ldots,T}$}
\STATE Sample action $a_t$ from $\eta$
\STATE Do action $a_t$ and observe $X_t$ and $Y_t$
\ENDFOR
\FOR{$a \in \calA$}
\STATE
\eq {
\hat \mu_a =  \frac{1}{T} \sum_{t=1}^T Y_t R_a(X_t)  \ind{R_a(X_t) \leq B_a}
}
\ENDFOR
\STATE {\bf return} $\hat a^*_T = \argmax_a \hat \mu_a$
\end{algorithmic}
\end{algorithm}
\end{minipage}
\end{wrapfigure}

@@

The second observation is that, in the parallel bandit setting, $m(\eta^*) \leq 2m(\boldsymbol{q})$. This is easy to see by letting $\eta_a = 1/2$ for $a = do()$ and $\eta_a = \ind{\P{X_i = j} \leq 1/m(\boldsymbol{q})} / 2m(\boldsymbol{q})$ for the actions corresponding to $do(X_i=j)$, and applying an argument like that for Proposition~\ref{pro:m-bound}. \ifsup 
The proof is in Section \ref{sec:m-equivelence}.
\else
The proof is in the supplementary materials.
\fi
